\pkg{Rgbp}: An \proglang{R} Package for Conjugate Gaussian, Poisson, and Binomial Hierarchical Modeling and Frequency Method Checking on Overdispersed Data

### 2015-07-27 jss/article_v4.tex
Carl revision

\DeclareMathOperator*{\argmax}{arg\,max}
%% almost as usual
\author{Hyungsuk Tak\\Harvard University \And 
             Joseph Kelly\\Google\And
             Carl Morris\\ Harvard University}
\title{\pkg{Rgbp}: An \proglang{R} Package for Conjugate Gaussian, Poisson, and Binomial Hierarchical Modeling and Frequency Method Checking on Overdispersed Data}
\linespread{2}
%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Hyungsuk Tak, Joseph Kelly, Carl Morris} %% comma-separated
\Plaintitle{Rgbp: An R Package for Conjugate Gaussian, Poisson, and Binomial Hierarchical Modeling and Frequency Method Checking on Overdispersed Data} %% without formatting
\Shorttitle{\pkg{Rgbp}: Hierarchical Modeling and Frequency Method Checking on Overdispersed Data} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an \proglang{R} package that utilizes  approximate Bayesian machinery to fit two-level conjugate hierarchical models on overdispersed Gaussian, Poisson, and Binomial data. The data that \pkg{Rgbp} assumes comprise of observed sufficient statistics for each random effect, such as sample means, possibly together with covariates of each group. The approximate Bayesian tool equipped with the adjustment for density maximization estimates random effects and hyper-parameters in the conjugate prior distributions. As for the Binomial model, the package has an option to draw their posterior samples via the acceptance-rejection method.  The main goal of \pkg{Rgb} is to produce  Bayesian interval estimates for the random effects that meet their nominal confidence levels.   For this purpose, we adopt unique improper hyper-prior distributions. \pkg{Rgbp} provides a tool to check quickly whether the resulting Bayesian interval estimates for the random effects achieve the nominal confidence levels via a repeated sampling coverage evaluation, which we call  ``frequency method checking."  }
% and evaluates repeated sampling frequency property of the resulting approximate Bayesian interval estimates for random effects
%It is found that such Bayesian-frequentist reconciliation allows \pkg{Rgbp} to have attributes desirable from both perspectives, working well in small samples and yielding good coverage probabilities for its interval estimates.
\Keywords{overdispersion, hierarchical model, adjustment for density maximization, acceptance-rejection method, repeated sampling coverage evaluation, frequency method checking, \proglang{R}}
\Plainkeywords{multilevel model, conjugate hierarchical generalized linear models, frequency method checking, coverage probability, shrinkage, R} 
%% without formatting
%% at least one keyword must be supplied

@@

\section[Introduction]{Introduction (v3; 2 July 2015)}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.
Gaussian, Poisson, or Binomial data from several independent groups sometimes have more variation than the assumed Gaussian, Poisson, or Binomial distributions of the first-level observed data. To account for this extra-variability, called overdispersion, a  two-level conjugate hierarchical model regards first-level mean parameters as random effects that come from a population-level conjugate prior distribution. The main  goal of our two-level conjugate modeling is to estimate these random effects for a comparison between groups.

@@

With an assumption of homogeneity within each group, the observed data are group-level aggregate data from independent $k$ groups, composed of sufficient statistics for the random effects without the population-level data. Specifically, the data for the Gaussian model comprise of each group's sample mean and its standard error, those for the Poisson model have each group's number of outcomes happening given some exposure for the Poisson data, and those for the Binomial model contain each group's number of successful outcomes out of total number of trials. The data may contain each group's covariate information. This type of  data is common for a biological analysis on litter data \citep{tamura1987stabilized}, a meta analysis on independent studies (Chapter 5 in \cite{gelman2014bayesian}), or small area estimation problems \citep{ghosh1994small, rao2003small}.

For these data with the assumption of homogeneity within each group, the two-level model that \pkg{Rgbp} assumes can be considered as a conjugate hierarchical generalized linear model (HGLM) \citep{lee1996hierarchical, hglm2006} where each random effect has a conjugate prior distribution. However, the HGLM mainly focuses on estimating regression coefficients to explore associations between covariates and observed data, while we are more concerned about estimating random effects.

\pkg{Rgbp} takes a Bayesian approach with our unique improper hyper-prior distributions on the parameters of the conjugate prior distribution, to produce  Bayesian interval   estimates for the random effects that achieve nominal confidence levels. The hyper-prior distributions lead to Stein's harmonic prior known to produce good repeated sampling coverage rates of the Bayesian interval estimates for random effects in a two-level Gaussian hierarchical model \citep{tang2011, morris2012, kelly2014advances}.  We apply an analog to Stein's harmonic prior to Poisson and Binomial hierarchical models. 
%The main goal of our Bayesian hierarchical models is to compare random effects between groups by their Bayesian interval estimates that have good frequency properties.

When it comes to fitting the model, \pkg{Rgbp} adopts the adjustment for density maximization \citep{carl1988, morris1997, tang2011} (ADM), a Pearson family approximation via maximization. For example, a Delta method is a Gaussian case of the ADM that uses a Normal distribution to obtain an approximate distribution of a function of a parameter with its MLE and negative Hessian at the MLE plugged-in. In this article, the ADM uses Beta distributions mainly to approximate the posterior distributions of shrinkage factors, functions of the second-level variance component, via maximization.  Using the approximate Beta posterior distributions of shrinkage factors, \pkg{Rgbp} estimates the first two (three for the Gaussian case) posterior moments of the random effects. Finally, \pkg{Rgbp} approximates the posterior distribution of each random effect by a skewed Normal distribution for a Gaussian case, a Gamma distribution for a Poisson case, and a Beta distribution for a Binomial case whose two parameters (three for the skewed Normal distribution) are matched to the previously estimated posterior moments of the random effects.

@@

For the Binomial  hierarchical model, \pkg{Rgbp} provides an option to draw independent posterior samples of all the model parameters including random effects via an acceptance-rejection method \citep{everson2000inference, tang2002fitting} instead of the approximate Bayesian tool.

In addition to fitting hierarchical models, \pkg{Rgbp} evaluates the repeated sampling coverage rates of the resulting Bayesian interval estimates for random effects \citep{morris1997, daniels1999prior, tang2002fitting, tang2011, morris2012}. It is a unique procedure that distinguishes \pkg{Rgbp}  from any other \proglang{R} packages for similar hierarchical modeling such as \pkg{hglm} \citep{hglm2010, ronnegaard2011hglm} for conjugate hierarchical generalized models. The evaluation procedure which we call ``frequency method checking'' uses a parametric bootstrapping  that generates mock data sets given the values of the hyper-parameters and estimates the coverage rates based on the generated mock data sets.

%:
The rest of this paper is organized as follows. We specify the Bayesian hierarchical models and discuss their posterior propriety in Section \ref{sec2}. In Section \ref{inference}, we explain the inferential models used to estimate the model parameters. We describe the estimation procedures including the ADM and the acceptance-rejection method in Section  \ref{sec3} and \ref{sec_accept}, respectively. We introduce the frequency method checking in Section \ref{sec4}.  We explain the usages of main functions in \pkg{Rgbp} in Section \ref{sec5}, and apply them to three examples in Section \ref{sec6}.

@@

\subsection[Normal-Normal]{Normal-Normal model for Gaussian data}
The following Normal-Normal hierarchical model (or the Gaussian model) assumed by \pkg{Rgbp} is useful when the group-level aggregate data from independent $k$ groups are continuous variables with known standard errors. The subscript \emph{j} below indicates the \emph{j}th group among \emph{k} groups in the dataset. For $j=1, 2, \ldots, k$, 
\begin{align}
y_{j}\vert \mu_{j} & \stackrel{\textrm{indep.}}{\sim} \textrm{Normal}(\mu_{j}, V_{j}),\label{normalobs}\\
\mu_{j}\vert \boldsymbol{\beta}, A & \stackrel{\textrm{indep.}}{\sim} \textrm{Normal}(\mu^E_{j}, A),\label{normalprior}
\end{align}

@@

%; sometimes we may be able to obtain  information about the mean of the population from previous studies.
%; in the limit of $r$ going to 0, the conjugate prior distribution gets flatter. 
\subsection[Poisson-Gamma]{Poisson-Gamma model for Poisson data}\label{poissonsubsec}
\pkg{Rgbp} can estimate a conjugate Poisson-Gamma hierarchical model (or the Poisson model) when the group-level aggregate data from independent $k$ groups comprise of non-negative count data without upper limitation. However, its usage is limited to the case where the expected random effect is known ($m=0$); we may obtain its information from the past studies or from experts. For $j=1, 2, \ldots, k$, 
\begin{align}
y_j\vert \lambda_j &\stackrel{\textrm{indep.}}{\sim}  \textrm{Poisson}(n_{j}\lambda_{j}),\\
\lambda_{j}\vert r &\stackrel{i.i.d.}{\sim} \textrm{Gamma}(r\lambda^E, r),
\end{align}
where $y_j$ is the number of events happening, $n_{j}$ is the exposure of group $j$, which is not necessarily an integer, $\lambda^E=\E(\lambda_j\vert r)$ is the known expected random effect ($m=0$), and $r$ is the unknown second-level variance component. The mean and variance of the conjugate Gamma prior distribution are $\lambda^E$ and $\lambda^E/r$, respectively.  \cite{albert1988computational} interprets $r$ as the amount of prior information as $n_{j}$ represents the amount of observed information because the uncertainty of the conjugate prior distribution increases as $r$ decreases and vice versa. The conditional posterior distribution of the random effect $\lambda_j$ for this Poisson model is
\begin{equation} \label{gammapost}
\lambda_j\vert r, \boldsymbol{y} \stackrel{\textrm{indep.}}{\sim}\textrm{Gamma}(r\lambda^E + n_j\bar{y}_{j},~ r + n_j),
\end{equation}  
where $\bar{y}_j\equiv y_j/n_j$. The mean and variance of the conditional posterior distribution are
\begin{equation}\label{gammapost_mean_var}
\lambda^\ast_j\equiv \E(\lambda_j\vert r, \boldsymbol{y} )=(1-B_{j})\bar{y}_{j} + B_{j}\lambda^E~~\textrm{and}~~\VAR(\lambda_j\vert r, \boldsymbol{y})=\frac{\lambda^\ast_j}{r+n_j}.
\end{equation}
where $B_{j}\equiv r / (r+n_{j})$ is the shrinkage factor of group $j$, the relative amount of information in the prior compared to the data. The conditional posterior mean is  a convex combination of $\bar{y}_{j}$ and  $\lambda^E$ weighted by $B_j$. If the conjugate prior distribution contains more information than the observed data have, \emph{i.e.}, ensemble sample size $r$ exceeds individual sample size $n_{j}$, then the posterior mean shrinks  towards the prior mean by more than 50\%.%, borrowing more information from the second-level distribution.

@@

\subsection[Binomial-Beta]{Binomial-Beta model for Binomial data}
\pkg{Rgbp} can fit a conjugate Binomial-Beta hierarchical model when the group-level aggregate data from independent $k$ groups are composed of each group's number of successes out of total number of trials. Unlike the Poisson model, the expected random effect is either known ($m=0$) or unknown ($m\ge1$). For $j=1, 2, \ldots, k$,
\begin{align}
y_{j} \vert p_{j} &\stackrel{\textrm{indep.}}{\sim} \textrm{Binomial}(n_{j}, ~p_{j}),\\
p_{j} \vert \boldsymbol{\beta}, r &\stackrel{\textrm{indep.}}{\sim} \textrm{Beta}(rp^E_j,~ r(1-p^E_j)),
\end{align}

>>>>

\DeclareMathOperator*{\argmax}{arg\,max}
%% almost as usual
\author{Hyungsuk Tak\\Harvard University \And 
             Joseph Kelly\\Google\And
             Carl Morris\\ Harvard University}
\title{\pkg{Rgbp}: An \proglang{R} Package for Conjugate Gaussian, Poisson, and Binomial Hierarchical Modeling and Frequency Method Checking on Overdispersed Data}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Hyungsuk Tak, Joseph Kelly, Carl Morris} %% comma-separated
\Plaintitle{Rgbp: An R Package for Conjugate Gaussian, Poisson, and Binomial Hierarchical Modeling and Frequency Method Checking on Overdispersed Data} %% without formatting
\Shorttitle{\pkg{Rgbp}: Hierarchical Modeling and Frequency Method Checking on Overdispersed Data} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an \proglang{R} package that utilizes  approximate Bayesian machinery to fit two-level conjugate hierarchical models on overdispersed Gaussian, Poisson, and Binomial data. The data that \pkg{Rgbp} assumes comprise of observed sufficient statistics for each random effect, such as sample means, possibly together with covariates of each group. The approximate Bayesian tool equipped with the adjustment for density maximization estimates random effects and hyper-parameters in the conjugate prior distributions. As for the Binomial model, the package has an option to draw their posterior samples via the acceptance-rejection method.  The main goal of \pkg{Rgb} is to produce  Bayesian interval estimates for the random effects that meet their nominal confidence levels.   For this purpose, we adopt unique improper hyper-prior distributions. \pkg{Rgbp} provides a tool to check quickly whether the resulting Bayesian interval estimates for the random effects achieve the nominal confidence levels via a repeated sampling coverage evaluation, which we call  ``frequency method checking."  }
% and evaluates repeated sampling frequency property of the resulting approximate Bayesian interval estimates for random effects
%It is found that such Bayesian-frequentist reconciliation allows \pkg{Rgbp} to have attributes desirable from both perspectives, working well in small samples and yielding good coverage probabilities for its interval estimates.
\Keywords{overdispersion, hierarchical model, adjustment for density maximization, acceptance-rejection method, repeated sampling coverage evaluation, frequency method checking, \proglang{R}}
\Plainkeywords{multilevel model, conjugate hierarchical generalized linear models, frequency method checking, coverage probability, shrinkage, R} 
%% without formatting
%% at least one keyword must be supplied

@@

\section[Introduction]{Introduction (v3; 27 July 2015)}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.
Gaussian, Poisson, or Binomial data from several independent groups sometimes have more variation than the assumed Gaussian, Poisson, or Binomial distributions of the first-level observed data. To account for this extra-variability, called overdispersion, a  two-level conjugate hierarchical model regards first-level mean parameters as random effects that come from a population-level conjugate prior distribution. The main  goal of our two-level conjugate modeling is to estimate these random effects for a comparison between groups.

@@

With an assumption of homogeneity within each group, the observed data are group-level aggregate data from $k$ independent  groups, composed of sufficient statistics for their $k$ random effects without the population-level data. Specifically, the data for the Gaussian model comprise of each group's sample mean and its standard error, those for the Poisson model use each group's outcome count and an exposure measure, those for the Binomial model use the number of each group's  successful outcomes together with the total number of trials. The data analyzed by \pkg{Rgbp} may incorporate each group's covariate information. These types of  data are common, e.g., as for a biological analysis on litter data \citep{tamura1987stabilized}, a meta analysis on independent studies (Chapter 5 in \cite{gelman2014bayesian}), or small area estimation problems \citep{ghosh1994small, rao2003small}.

For such data, assuming homogeneity within each group, \pkg{Rgbp}'s two-level model may be viewed as a conjugate hierarchical generalized linear model (HGLM) \citep{lee1996hierarchical, hglm2006} where each random effect has a conjugate prior distribution.    However, the HGLM focuses on estimating regression coefficients to explore associations between covariates and observed data.   While \pkg{Rgbp} does that too, its emphasis concerns making valid point and interval estimates for the $k$ random effects.

\pkg{Rgbp} employs a Bayesian approach with our unique improper hyper-prior distributions placed on the second-level parameters of the conjugate prior distribution. The hyper-prior is chosen to produce Bayesian interval   estimates for the random effects that achieve nominal confidence levels. The hyper-prior distribution used in \pkg{Rgbp} is related to Stein's harmonic prior for two-level Gaussian models where it has been seen to produce good repeated sampling coverage rates for the Bayesian interval estimates for the $k$ random effects \citep{tang2011, morris2012, kelly2014advances}.  This prior, as adapted further here for \pkg{Rgbp}'s Poisson and Binomial hierarchical models, has been seen for every data set tested thus far to meet (or exceed) the pre-assigned confidence level. 
%The main goal of our Bayesian hierarchical models is to compare random effects between groups by their Bayesian interval estimates that have good frequency properties.

For fitting the hierarchical model, \pkg{Rgbp} uses ADM, i.e. ``adjustment for density maximization'' \citep{carl1988, morris1997, tang2011}.   ADM approximates a posterior density or a likelihood function by fitting a pre-specified (one dimensional) Pearson family, using the first two derivatives of the given function.  When the Normal distribution is the chosen Pearson family, ADM produces exactly the same approximation as maximum likelihood estimation (MLE) does for fitting a Normal curve.   Because shrinkage factors lie in [0,1], \pkg{Rgbp} approximates shrinkage factor distributions with Beta distributions because they are constrained to [0,1], unlike the Normal.

\pkg{Rgbp} estimates the first two (actually, three in the Gaussian case) posterior moments of each random effect.   Finally, \pkg{Rgbp} approximates the posterior distribution of each random effect by a skewed Normal distribution for Gaussian case, by a Gamma distribution for the Poisson case, and by a Beta distribution for the Binomial case.

@@

For the Binomial  hierarchical model, \pkg{Rgbp} provides an option to draw independent posterior samples of all the model parameters, including random effects, via an acceptance-rejection method \citep{everson2000inference, tang2002fitting}.

In addition to fitting the hierarchical models, \pkg{Rgbp} evaluates the repeated sampling coverage rates of the resulting Bayesian interval estimates for random effects \citep{morris1997, daniels1999prior, tang2002fitting, tang2011, morris2012}. This is a unique procedure that distinguishes \pkg{Rgbp}  from any other \proglang{R} packages for similar hierarchical modeling such as \pkg{hglm} \citep{hglm2010, ronnegaard2011hglm} for conjugate hierarchical generalized models. The evaluation procedure which we use for this process of ``frequency method checking'' uses a parametric bootstrapping  method that many times generates mock data sets given the fitted values of the estimated hyper-parameters and uses the mock data to estimate the coverage rates.

%:
The rest of this paper is organized as follows. We specify the Bayesian hierarchical models and discuss their posterior propriety in Section \ref{sec2}. In Section \ref{inference}, we explain the inferential models used to estimate the model parameters. We describe the estimation procedures including  ADM and the acceptance-rejection method in Section  \ref{sec3} and \ref{sec_accept}, respectively. We introduce  frequency method checking techniques in Section \ref{sec4}.  We explain the usages of \pkg{Rgbp}'s main functions in \pkg{Rgbp} in Section \ref{sec5}, and apply them to three data examples in Section \ref{sec6}.

@@

\subsection[Normal-Normal]{Normal-Normal model for Gaussian data}
The following Normal-Normal hierarchical model (or the Gaussian model) assumed by \pkg{Rgbp} is useful when the group-level aggregate data from $k$ independent  groups are continuous variables with known standard errors. The subscript \emph{j} below indicates the \emph{j}th group among \emph{k} groups in the dataset. For $j=1, 2, \ldots, k$, 
\begin{align}
y_{j}\vert \mu_{j} & \stackrel{\textrm{indep.}}{\sim} \textrm{Normal}(\mu_{j}, V_{j}),\label{normalobs}\\
\mu_{j}\vert \boldsymbol{\beta}, A & \stackrel{\textrm{indep.}}{\sim} \textrm{Normal}(\mu^E_{j}, A),\label{normalprior}
\end{align}

@@

%; sometimes we may be able to obtain  information about the mean of the population from previous studies.
%; in the limit of $r$ going to 0, the conjugate prior distribution gets flatter. 
\subsection[Poisson-Gamma]{Poisson-Gamma model for Poisson data}\label{poissonsubsec}
\pkg{Rgbp} can estimate a conjugate Poisson-Gamma hierarchical model (or the Poisson model) when the group-level aggregate data from $k$ independent  groups comprise of non-negative count data without upper limitation. However, its usage is limited to the case where the expected random effect is known ($m=0$); we may obtain its information from the past studies or from experts. For $j=1, 2, \ldots, k$, 
\begin{align}
y_j\vert \lambda_j &\stackrel{\textrm{indep.}}{\sim}  \textrm{Poisson}(n_{j}\lambda_{j}),\\
\lambda_{j}\vert r &\stackrel{i.i.d.}{\sim} \textrm{Gamma}(r\lambda^E, r),
\end{align}
where $y_j$ is the number of events happening, $n_{j}$ is the exposure of group $j$, which is not necessarily an integer, $\lambda^E=\E(\lambda_j\vert r)$ is the known expected random effect ($m=0$), and $r$ is the unknown second-level variance component. The mean and variance of the conjugate Gamma prior distribution are $\lambda^E$ and $\lambda^E/r$, respectively.  \cite{albert1988computational} interprets $r$ as the amount of prior information as $n_{j}$ represents the amount of observed information because the uncertainty of the conjugate prior distribution increases as $r$ decreases and vice versa. The conditional posterior distribution of the random effect $\lambda_j$ for this Poisson model is
\begin{equation} \label{gammapost}
\lambda_j\vert r, \boldsymbol{y} \stackrel{\textrm{indep.}}{\sim}\textrm{Gamma}(r\lambda^E + n_j\bar{y}_{j},~ r + n_j),
\end{equation}  
where $\bar{y}_j\equiv y_j/n_j$. The mean and variance of the conditional posterior distribution are
\begin{equation}\label{gammapost_mean_var}
\lambda^\ast_j\equiv \E(\lambda_j\vert r, \boldsymbol{y} )=(1-B_{j})\bar{y}_{j} + B_{j}\lambda^E~~\textrm{and}~~\VAR(\lambda_j\vert r, \boldsymbol{y})=\frac{\lambda^\ast_j}{r+n_j}.
\end{equation}
where $B_{j}\equiv r / (r+n_{j})$ is the shrinkage factor of group $j$, the relative amount of information in the prior compared to the data. The conditional posterior mean is  a convex combination of $\bar{y}_{j}$ and  $\lambda^E$ weighted by $B_j$. If the conjugate prior distribution contains more information than the observed data have, \emph{i.e.}, ensemble sample size $r$ exceeds individual sample size $n_{j}$, then the posterior mean shrinks  towards the prior mean by more than 50\%.%, borrowing more information from the second-level distribution.

@@

\subsection[Binomial-Beta]{Binomial-Beta model for Binomial data}
\pkg{Rgbp} can fit a conjugate Binomial-Beta hierarchical model when the group-level aggregate data from $k$ independent  groups are composed of each group's number of successes out of total number of trials. Unlike the Poisson model, the expected random effect is either known ($m=0$) or unknown ($m\ge1$). For $j=1, 2, \ldots, k$,
\begin{align}
y_{j} \vert p_{j} &\stackrel{\textrm{indep.}}{\sim} \textrm{Binomial}(n_{j}, ~p_{j}),\\
p_{j} \vert \boldsymbol{\beta}, r &\stackrel{\textrm{indep.}}{\sim} \textrm{Beta}(rp^E_j,~ r(1-p^E_j)),
\end{align}
